\Title

### 2013-08-08 response_to_reviewers.txt

>These are interesting results and the manuscript is clearly written. The second
>observation is not novel, but underappreciated, whereas this is the first
>concrete calculation of the average selective disadvantage of silent mutations
>in HIV taking draft into account. As such, it is necessary for this work to
>establish proper error estimates so that the community can make valid use of
>the results.  With this in mind, the manuscript can be strengthened in the
>following respects:

We thank the reviewer for the constructive comments, which have helped
us to improve the manuscript. We have resorted to a more sophisticated
sampling scheme for parameters in the computer simulations and now scan
a larger parameter space. All our conclusions remain valid, but are more
reliable now. We detail these improvements in our point to point
response below.

##############################################################################
>1) The modeling generation time of 1 day is odd.  It takes about a day for the
>virion to integrate and start transcription.  One usually takes 2 days as the
>mean generation time, but there is individual variation. See next point.

We agree that one generation in HIV takes closer to 2 days than 1
day. Most escape rates and selection coefficients, however, are more
readily measured in days and we have chosen to model HIV evolution with
generation of one day for simplicity. Factor of 2 changes in the
generation have next to now influence on the dynamics and only
renormalize genetic drift. Genetic drift is not a dominant force in HIV
evolution and the population size, the main determinant of drift, is not
known to this accuracy anyway. We have repeated a subset of the
simulations with a generation time of one day, finding no
difference. IS THIS SHOWN SOMEWHERE?

@@

Indeed. Our previous simulations where not limited by coverage of the
parameter space. Now that we cover a much larger (higher dimensional)
space, we did implement a Latin hypercube sampling scheme. We thank the
reviewer for this very useful suggestion. WE MIGHT WANT TO SHOW THE RANK
CORRELATION MATRIX BETWEEN THE DIFFERENT VARIABLES.

##############################################################################
>5) The actual dynamics of HIV in the body is presumably more complex: it is
>likely that the virions are produced in large bursts of virtually genetically
>identical virions most of which are then cleared by the body fast on the
>time-scale of HIV generations. There is no discussion whether this leads to
>increased stochasticity in the dynamics, nor whether this can influence the
>quantitative evaluations presented here. It is possible that the authors mean
>to absorb these into some kind of reduced effective population size, but a
>short discussion would be beneficial.

We agree with reviewer that HIV evolution is bursty, but we would like
to mention that actual number of cells that get infected every day is
around 10^7 and hence much larger than we are able to simulate. The
population size in our simulation can therefore be viewed as an
effective population size. We have investigated the sensitivity of our
results to changes in the population size and find that our conclusions
are insensitive to the population size as long as the product of
population size and mutation rate is not very much smaller than 1.

We have added a discussion of our choice of population size to the
methods section next the part where the other parameter choices are
discussed.

@@

##############################################################################
>6) In the discussions, the rise of synonymous mutations is attributed to draft,
>but the fall to selective effects. In general, some of the fall might also be
>due to overlapping selective sweeps: the manuscript does not discuss this
>possibility.

Yes, much of the stochastic component in the dynamics of the frequencies
of synonymous mutations will be due to draft. The consistent loss of
variants, conditional on them being at high frequency earlier, implies
selection against these variants. We have added a sentence to the
discussion to make this point clear.

@@

##############################################################################
>7) The observed neutral fixation pattern of the nonsynonymous mutations is
>mentioned, and two explanations have been proposed to duet of increased
>diversity over synonymous mutations with no increase of fixation over neutral
>expectations.  Neither of the explanations, however, are plausible (at least,
>without argument) absent experimental evidence not mentioned in the text. It
>would perhaps be better to either strengthen this discussion with citation to
>literature, or to weaken it by mentioning these as mere possibilities amongst
>others.

We agree with the reviewer that the both hypothesis we put forward to
explain the fixation pattern of non-synonymous mutations are somewhat
tentative. Qualitatively, however, the processes we discuss have been
observed and we do cite experimental evidence. It is common for escape
to happen in multiple equivalent ways and this has been well documented
fro early CTL escape and early nAb escape (Wei et al, Nature, 2003. Bar
et al, PPath, 2012). Data for chronic infection is less conclusive and
selection pressures are generally smaller, but turnover at targeted
epitopes and the replacement of variants that where previously at high
frequencies was observed 3y into the infection in Wei et al, Nature, 2003.
Whether this fall in frequency of some non-synonymous variants happens
because they are replaced by a less costly escape variant or because the
escape variant itself is targeted is hard to disentangle. We think it is
reasonable that both of these effects contribute. Whether they are
sufficient to produce the observed fixation patterns of non-synonymous
mutations is largely an empirical question. We simply explored using
numerical experiments whether these effects push the curve in the
"right" direction.

### 2013-08-08 response_to_reviewers.txt

>These are interesting results and the manuscript is clearly written. The second
>observation is not novel, but underappreciated, whereas this is the first
>concrete calculation of the average selective disadvantage of silent mutations
>in HIV taking draft into account. As such, it is necessary for this work to
>establish proper error estimates so that the community can make valid use of
>the results.  With this in mind, the manuscript can be strengthened in the
>following respects:

We thank the reviewer for the constructive comments, which have helped
us to improve the manuscript. We have resorted to a more sophisticated
sampling scheme for parameters in the computer simulations and now scan
a larger parameter space. All our conclusions remain valid, but are more
reliable now. We detail these improvements in our point to point
response below.

##############################################################################
>1) The modeling generation time of 1 day is odd.  It takes about a day for the
>virion to integrate and start transcription.  One usually takes 2 days as the
>mean generation time, but there is individual variation. See next point.

We agree that one generation in HIV takes closer to 2 days than 1
day. Most escape rates and selection coefficients, however, are more
readily measured in days and we have chosen to model HIV evolution with
generation of one day for simplicity. Factor of 2 changes in the
generation have little influence on the dynamics and only
renormalize genetic drift. Genetic drift is not a dominant force in HIV
evolution and the population size, the main determinant of drift, is not
known to this accuracy anyway. We have repeated a subset of the
simulations with a generation time of one day, finding no
difference.

IS THIS SHOWN SOMEWHERE? (Fabio is redoing the simulations)

@@

##############################################################################
>2) Related to the above: the simulation does not address how sensitive the
>results are to varying the parameters (in particular, the generation time, the
>outcrossing rate, the mutation rate, and the assigned deleterious fitness of
>the "non-synonymous" changes).  In the absence of described theoretical
>analysis shaping our expectations, I strongly feel that this sensitivity
>analysis needs to be part of the results, and folded into the error bars.

We have greatly extended the parameter space that we survey and now test
different values for the mutation rate, the recombination rate, the
generation time, the fitness effects, and the population size. To
accomodate this high dimensional data, we have changed figure 4. The
figure now shows the joint distribution of the deleterious effect and
the fraction of synonymous mutations that are deleterious that are
compatible with observations as a heat map. The marginal distribution of
the individual parameters consistent with the data are shown in the supplement.

The distribution of allowed deleterious is sharply peaked around
0.02. DISCUSS THE DIVERSITY ISSUES.

@@

##############################################################################
>3) The reasonableness of the ranges for the various parameters should be argued
>with citation to existing literature.

We have added a discussion of the parameter ranges to the methods
section.

##############################################################################
>4) I am a bit surprised that the authors chose to do random sampling in their
>multidimensional parameter space.  One would have expected a Latin hypercube
>sampling to provide a better coverage.

Indeed. Our previous simulations were not limited by coverage of the
parameter space. Now that we cover a much larger (higher dimensional)
space, we did implement a Latin hypercube sampling scheme. We thank the
reviewer for this very useful suggestion. WE MIGHT WANT TO SHOW THE RANK
CORRELATION MATRIX BETWEEN THE DIFFERENT VARIABLES.

@@

We have added a discussion of our choice of population size to the
methods section next the part where the other parameter choices are
discussed.

##############################################################################
>6) In the discussions, the rise of synonymous mutations is attributed to draft,
>but the fall to selective effects. In general, some of the fall might also be
>due to overlapping selective sweeps: the manuscript does not discuss this
>possibility.

Yes, much of the stochastic component in the dynamics of the frequencies
of synonymous mutations will be due to draft. The consistent loss of
variants, conditional on them being at high frequency earlier, implies
selection against these variants. We have added a sentence to the
discussion to make this point clear.

Fabio: I DO NOT THINK THAT ADRESSES THE QUESTION ("IN GENERAL" --> "BUT").
MAYBE WE COULD STATE SOMETHING LIKE: "IN ABSENCE OF DELETERIOUS EFFECTS, THE
FIXATION PROBABILITY OF HITCHHIKERS IS NOT AFFECTED BY DRAFT".

### 2013-08-08 synmut.tex
few comments on response and manuscript

We find selection against synonymous substitutions despite the
fact that the corresponding sites are not strongly conserved in
cross sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of these structures
changes over time, selection seems to maintain them, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes it hard to find the 
compensatory mutations that would restore base pairing in the
longitudinal data; in fact, the exact base pairing pattern is
most likely different than in the reference sequence. Nevertheless, the
importance of secondary structure underscores the realization that viral
fitness is a complicated function (fitness landscape) in the
high-dimensional space of possible genotypes
\citep{ferguson_translating_2013}. We show, that this landscape is shaped not
only by interactions between amino-acids, but also by ubiquitous
interactions between nucleotides in RNA structures.

>>>>

We find selection against synonymous substitutions despite the
fact that the corresponding sites are not strongly conserved in
cross sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of these structures
changes over time, selection seems to maintain them, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes it hard to find the 
compensatory mutations that would restore base pairing in the
longitudinal data; in fact, the exact base pairing pattern is
most likely different than in the reference sequence. Nevertheless, the
importance of secondary structure underscores the realization that viral
fitness is a complicated function (fitness landscape) in the
high-dimensional space of possible genotypes
\citep{ferguson_translating_2013}. We show, that this landscape is shaped not
only by interactions between amino acids, but also by ubiquitous
interactions between nucleotides in RNA structures.
